---
title: "DATA202/472 - 2024 - Assignment 1"
author: "Jackson Rakena"
date: "2024-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=TRUE)
```


## Q1
a. Numerical continuous, class `numeric`, type `double`
b. Categorical logical, class `logical`, type `logical`
c. Numerical integer, class `numeric`, type `integer`
d. Categorical nominal, class `character`, type `character`
e. Categorical ordinal, class `ordered factor`, type `integer`


## Q2
```{r}
y <- c("Earth"=0, "Water"=-3, "Fire"=2)
```
a.
```{r}
y[y==2]
```
b.
```{r}
names(y[y==2])
```
c.
```{r}
a <- c(3, 4, 1)
z <- array(c(0,3,0,9,2,1), dim=c(2,3))
print(a)
print(z)
```
d.
```{r}
max(a)
```


## Q3
a.
```{r}
standardize_column <- function(mat, col) {
  mean <- mean(mat[, col], na.rm=TRUE)
  sd <- sd(mat[, col], na.rm=TRUE)
  
  return((mat[, col] - mean) / sd)
}
```
b.
```{r}
mat <- matrix(c(12,1,4,3,NA,9,NA,5,5), nrow=3)
print(standardize_column(mat, 2))
```
c.
```{r}
standardize_column <- function(mat, col) {
  if (col > ncol(mat) || !is.vector(mat[,col])) {
    stop("col must be a valid index of mat")
  }
  mean <- mean(mat[, col], na.rm=TRUE)
  sd <- sd(mat[, col], na.rm=TRUE)
  
  return((mat[, col] - mean) / sd)
}
```
d.
```{r, error=TRUE}
print(standardize_column(mat, 0.3))
print(standardize_column(mat, -1))
print(standardize_column(mat, 1))
print(standardize_column(mat, 4))
```

## Q4
a.
```{r}
stars <- read.csv("star_dataset_deepu_kaggle.csv", header=TRUE)
```
b.
```{r}
stars[1:2,]
```
c.
```{r}
nrow(stars)
ncol(stars)
```
d.
```{r}
subset_stars <- stars[stars$Star.color %in% c("Orange","White","Blue"),]
head(subset_stars)
```
e.
```{r}
is.factor(subset_stars$Star.color)
```
f.
```{r}
subset_stars$star_colours <- factor(subset_stars$Star.color, levels = c("Blue", "White", "Orange"))
```
g.
```{r}
barplot(table(subset_stars$star_colours), main = "Counts of star colours", xlab = "Star Color", ylab = "Frequency", col = c("blue", "white", "orange"))
```


## Q5
a. This statement sets the value of `x` to `y`.
b. This statement checks the equality of `x` and `y`, i.e. that all elements of `x` are identical to `y`.
c. This statement also sets the value of `x` to `y`.
d. This checks that `x` is less than the inverse of `y`.
e. This takes each element of `x` and puts it the power of the the negative value of the element of `y` in the same position.


## Q6
```{r}
library(palmerpenguins)
```
a.
```{r}
hist(penguins$body_mass_g, main="Distribution of body masses",
     xlab="Body mass (g)", col="lightblue")
```

b.
```{r}
plot(penguins$flipper_length_mm, penguins$bill_length_mm,
     xlab="Flipper length (mm)", ylab="Bill length (mm)",
     main="Flipper lengths by bill lengths")
```

c.
```{r}
barplot(table(penguins$species, penguins$island),
        main = "Counts of Penguins by Species and Island",
        xlab = "Species",
        ylab = "Count",
        legend.text = TRUE,
        col = c("lightblue", "lightgreen", "lightpink"),
        beside = TRUE)
```

d.
```{r}
boxplot(penguins$body_mass_g ~ penguins$sex,
        main = "Boxplot of Body Mass by Sex",
        xlab = "Sex",
        ylab = "Body Mass (g)",
        col = c("lightblue", "lightblue"),
        border = "black",
        notch = TRUE)
```


## Q7
Fixed version:
```{r}
# Define x
x <- c(3,0,5,6,NA,3,1) 
# Store the first five entries of x as a new object called y
y <- x[1:5]
# Calculate the maximum value of y after excluding NA entries
max_y <- max(y, na.rm=TRUE)
# Fetch the entries of y that are bigger than 2
y[y > 2]
```

Fixes:

1. Line 2: To define a variable we need to use `<-` instead of `<=`
2. Line 4: We need to use `1:5` to select the first five elements
3. Line 6: We need to use `na.rm` instead of `na.remove` to remove NA elements
4. Line 8: We need to use `>` (bigger than) instead of `>=` (bigger than or equal to) to get elements bigger than 2